---
title: "CSR Opioids"
author: "Catherine DeJager, Sarah Gilbert, and Kaitlyn Westra"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    fig_height: 2.2
    fig_width: 4
  pdf_document:
    fig_height: 2.2
    fig_width: 4
  word_document:
    fig_height: 2.2
    fig_width: 4
---

```{r, setup, include = FALSE}
# load packages that are going to be used
require(tidyverse)   # this loads mosaic, ggformula, etc. too
require(ggformula)
require(mosaic)
require(plotly)
require(car)
require(MuMIn)

# Some customization.  You can alter or delete as desired (if you know what you are doing).

theme_set(theme_bw(base_size=12))     # change theme for ggplot2/ggformula

knitr::opts_chunk$set(
  echo = TRUE,      # for homework, always show R code (this is the default)
  tidy = FALSE,     # display code as typed (rather than reformatted)
  size = "small",   # slightly smaller font for code
  message = FALSE, warning = FALSE) # don't print warnings or messages in compiled document. So you MUST check them in RStudio!
```

### Read in the data

```{r}
load(file = "data/tedsd_2017.RData")
```

```{R}
glimpse(tedsd_puf_2017)
```

```{r}
tedsd_MI2017 <- filter(tedsd_puf_2017, STFIPS == 26)
```

```{r}
tedsd_MI2017 <- tedsd_MI2017 %>% 
  mutate(OPIFLG = HERFLG==1 | METHFLG==1 | OPSYNFLG==1, # MULTIOPI = (HERFLG==1 & METHFLG==1) | (HERFLG==1 & OPSYNFLG==1) | (METHFLG==1 & OPSYNFLG==1),
         HER1 = SUB1==5, HER2 = SUB2==5, HER3 = SUB3==5, METH1 = SUB1==6, METH2 = SUB2==6, METH3 = SUB3==6, OPSYN1 = SUB1==7, OPSYN2 = SUB1==7, OPSYN3 = SUB3==7)
tedsd_MI2017 <- tedsd_MI2017 %>% mutate(OPI1 = HER1|METH1|OPSYN1, OPI2 = HER2|METH2|OPSYN2, OPI3 = HER3|METH3|OPSYN3)
tedsd_MI2017 <- tedsd_MI2017 %>% mutate(MULTIOPI = (OPI1 & OPI2) | (OPI1 & OPI3) | (OPI2 & OPI3)) # FRSTOPI = ?
```

```{r}
remove_nas <- function(df) {
  M <- as.matrix(df)
  M[M == -9] <- NA
  result <- as.data.frame(M)
  result <- result %>% na.omit()
}
```

### Research Questions

1. What variables predict whether a person entering treatment uses opioids or not?
2. What variables predict whether a person entering treatment uses more than one kind of opioid?
3. What variables predict the length of stay for a person who enters treatment for opioid use?

#### 1. What variables predict whether a person entering treatment uses opioids or not?

```{r}
tedsd_MI2017_notna <- tedsd_MI2017 %>% select(HERFLG, METHFLG, OPSYNFLG, OPIFLG, HER1, HER2, HER3, METH1, METH2, METH3, OPSYN1, OPSYN2, OPSYN3, OPI1, OPI2, OPI3, MULTIOPI, AGE, GENDER, RACE, MARSTAT, EDUC, EMPLOY, LIVARAG, ARRESTS, NOPRIOR, PSYPROB, FREQ_ATND_SELF_HELP, LOS, PSOURCE)
tedsd_MI2017_notna <- tedsd_MI2017_notna %>% remove_nas()
glimpse(tedsd_MI2017_notna)
```

```{r}
tedsd_MI2017_notna <- tedsd_MI2017_notna %>% mutate(
    AGE = factor(AGE, labels = c("12-14", "15-17", "18-20", "21-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-64", "65+")),
    GENDER = factor(as.numeric(GENDER==2.0), labels = c("M", "F")),
    RACE = factor(RACE, labels = c("AK Native", "American Indian", "Black", "White", "Asian", "Other", "Multiple", "HI/other pacific islander")),
    MARSTAT = factor(MARSTAT, labels = c("Never", "Married", "Separated", "Divorced/widowed")),
    EDUC = factor(EDUC, labels = c("<=8", "9-11", "12 or GED", "13-15", "16+")),
    EMPLOY = factor(EMPLOY, labels = c("Full-time", "Part-time", "Unemployed", "Not in labor force")),
    LIVARAG = factor(LIVARAG, labels = c("Homeless", "Dependent", "Independent")),
    ARRESTS = factor(ARRESTS, labels = c("0", "1", "2+")),
    NOPRIOR = factor(NOPRIOR, labels = c("0", "1+")),
    PSYPROB = factor(as.numeric(PSYPROB == 1)),
    FREQ_ATND_SELF_HELP = factor(FREQ_ATND_SELF_HELP, labels = c("0", "1-3", "4-7", "8-30")),
    LOS = factor(LOS, labels = c(seq(from = 1, to = 30), "31-45", "46-60", "61-90", "91-120", "121-180", "181-365", ">365")),
    PSOURCE = factor(PSOURCE, labels = c("Individual", "Alcohol/drug use care provider", "Other health care provider", "School", "Employer/EAP", "Other community", "Court"))
    )
tedsd_MI2017_notna <- tedsd_MI2017_notna %>% mutate_if(is.double, as.factor)
glimpse(tedsd_MI2017_notna)
```

```{r}
tedsd_MI2017_opi <- tedsd_MI2017_notna %>% filter(OPIFLG == 1)
glimpse(tedsd_MI2017_opi)
```

```{r}
percents_graphs_tables <- function(form, ttl, my_df, horiz=FALSE) {
  if (horiz) {
    print(gf_percentsh(form, title=ttl, data=my_df))
  }
  else {
    print(gf_percents(form, title=ttl, data=my_df))
  }
  tally(form, format='percent', data=my_df)
}
```

```{r}
percents_graphs_tables(~HERFLG, "Heroin use in opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~METHFLG, "Non-prescription Methadone use in opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~OPSYNFLG, "Other Opiates and Synthetics use in opioid users", tedsd_MI2017_opi)
```

```{r}
opi_mod <- glm(OPIFLG ~ AGE + GENDER + MARSTAT + EDUC + EMPLOY + LIVARAG + ARRESTS + NOPRIOR + FREQ_ATND_SELF_HELP, na.action = na.fail, family=binomial(link = "logit"), data=tedsd_MI2017_notna)
```

```{r}
summary(opi_mod)
```

```{r}
vif(opi_mod)
```

```{r}
opi_mod <- glm(OPIFLG ~ RACE + GENDER + MARSTAT + EDUC + EMPLOY + LIVARAG + ARRESTS + NOPRIOR + FREQ_ATND_SELF_HELP, na.action = na.fail, family=binomial(link = "logit"), data=tedsd_MI2017_notna)
summary(opi_mod)
```

```{r}
d_res <- dredge(opi_mod, rank = "AIC")
head(d_res, 10)
```

```{r}
b_res <- dredge(opi_mod, rank = "BIC")
head(b_res, 10)
```

```{r}
interaction_mod <- glm(OPIFLG ~ GENDER + MARSTAT + GENDER*MARSTAT + EDUC + EMPLOY + GENDER*EMPLOY + EDUC*EMPLOY + LIVARAG + GENDER*LIVARAG + LIVARAG*EDUC + LIVARAG*EMPLOY + ARRESTS + NOPRIOR + GENDER*NOPRIOR + FREQ_ATND_SELF_HELP, na.action = na.fail, family=binomial(link = "logit"), data=tedsd_MI2017_notna)
summary(interaction_mod)
```

```{r}
interaction_mod <- glm(OPIFLG ~ GENDER + MARSTAT + GENDER*MARSTAT + GENDER*EMPLOY + LIVARAG + GENDER*LIVARAG + LIVARAG*EDUC + LIVARAG*EMPLOY + NOPRIOR + FREQ_ATND_SELF_HELP, na.action = na.fail, family=binomial(link = "logit"), data=tedsd_MI2017_notna)
summary(interaction_mod)
```

```{r}
vif(interaction_mod)
```

```{r}
a_interaction_res <- dredge(interaction_mod)
head(a_interaction_res, 10)
```

```{r}
percents_graphs_tables(~OPIFLG | AGE, "Opioid use by age at admission", tedsd_MI2017_notna)
percents_graphs_tables(~AGE, "Age at admission of Opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~OPIFLG | GENDER, "Opioid use by gender", tedsd_MI2017_notna)
percents_graphs_tables(~GENDER, "Gender of Opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~OPIFLG | MARSTAT, "Opioid use by marital status", tedsd_MI2017_notna)
percents_graphs_tables(~MARSTAT, "Marital status of Opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~OPIFLG | EDUC, "Opioid use by education level", tedsd_MI2017_notna)
percents_graphs_tables(~EDUC, "Education level of Opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~OPIFLG | EMPLOY, "Opioid use by employment at admission", tedsd_MI2017_notna)
percents_graphs_tables(~EMPLOY, "Employment at admission of opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~OPIFLG | LIVARAG, "Opioid use by age at admission", tedsd_MI2017_notna)
percents_graphs_tables(~LIVARAG, "Age at admission of opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~OPIFLG | ARRESTS, "Opioid use by number of arrests in the 30 days prior to admission", tedsd_MI2017_notna)
percents_graphs_tables(~ARRESTS, "Number of arrests in the 30 days prior to admission of opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~OPIFLG | NOPRIOR, "Opioid use by number of prior admissions", tedsd_MI2017_notna)
percents_graphs_tables(~NOPRIOR, "Number of prior admissions for opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~OPIFLG | PSYPROB, "Opioid use by presence of co-occuring mental and substance use disorders", tedsd_MI2017_notna)
percents_graphs_tables(~PSYPROB, "Presence of co-occuring mental and substance use disorders of opioid users", tedsd_MI2017_opi)
percents_graphs_tables(~OPIFLG | FREQ_ATND_SELF_HELP, "Opioid use by frequency of attending self-help programs in the past month", tedsd_MI2017_notna)
percents_graphs_tables(~FREQ_ATND_SELF_HELP, "Frequency of attending self-help programs in the past month of Opioid users", tedsd_MI2017_opi)
```

#### 2. What variables predict whether a person entering treatment uses more than one kind of opioid?

```{r}
percents_graphs_tables(~MULTIOPI | HERFLG, "Multiple opioid use by heroin use", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | METHFLG, "Multiple opioid use by non-prescription methadone use", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | OPSYNFLG, "Multiple opioid use by other opiates and synthetics use", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | AGE, "Multiple opioid use by age at admission", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | GENDER, "Multiple opioid use by gender", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | MARSTAT, "Multiple opioid use by marital status", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | EDUC, "Multiple opioid use by education level", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | EMPLOY, "Multiple opioid use by employment at admission", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | LIVARAG, "Multiple opioid use by age at admission", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | ARRESTS, "Multiple opioid use by number of arrests in the 30 days prior to admission", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | LIVARAG, "Multiple opioid use by living arrangement at admission", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | NOPRIOR, "Multiple opioid use by number of prior admissions", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | PSYPROB, "Multiple opioid use by presence of co-occuring mental and substance use disorders", tedsd_MI2017_notna)
percents_graphs_tables(~MULTIOPI | FREQ_ATND_SELF_HELP, "Multiple opioid use by frequency of attending self-help programs in the past month", tedsd_MI2017_notna)
```

#### 3. What variables predict the length of stay for a person who enters treatment for opioid use?

```{r}
percents_graphs_tables(~LOS | OPIFLG, "Distribution of length of stay by opioid use", tedsd_MI2017_notna, horiz=TRUE)
percents_graphs_tables(~LOS, "Distribution of length of stay for opioid users", tedsd_MI2017_opi, horiz=TRUE)
```

```{r}
tedsd_MI2017_los <- tedsd_MI2017 %>% select(HERFLG, METHFLG, OPSYNFLG, OPIFLG, HER1, HER2, HER3, METH1, METH2, METH3, OPSYN1, OPSYN2, OPSYN3, OPI1, OPI2, OPI3, MULTIOPI, AGE, GENDER, RACE, MARSTAT, EDUC, EMPLOY, LIVARAG, ARRESTS_D, SERVICES, METHUSE, DAYWAIT, REASON, PSOURCE, NOPRIOR, PSYPROB, FREQ_ATND_SELF_HELP, LOS) %>% filter(OPIFLG == 1)
glimpse(tedsd_MI2017_los)
```

```{r}
tedsd_MI2017_los <- tedsd_MI2017_los %>% remove_nas()
glimpse(tedsd_MI2017_los)
```

```{r}
tally(~AGE, data=tedsd_MI2017_notna)
tally(~AGE, data=tedsd_MI2017_opi)
tally(~SERVICES, data=tedsd_MI2017_los)
```

```{r}
tedsd_MI2017_los <- tedsd_MI2017_los %>% mutate(
    AGE = factor(AGE, labels = c("15-17", "18-20", "21-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-64", "65+")),
    GENDER = factor(as.numeric(GENDER==2.0), labels = c("M", "F")),
    RACE = factor(RACE, labels = c("AK Native", "American Indian", "Black", "White", "Asian", "Other", "Multiple", "HI/other pacific islander")),
    MARSTAT = factor(MARSTAT, labels = c("Never", "Married", "Separated", "Divorced/widowed")),
    EDUC = factor(EDUC, labels = c("<=8", "9-11", "12 or GED", "13-15", "16+")),
    EMPLOY = factor(EMPLOY, labels = c("Full-time", "Part-time", "Unemployed", "Not in labor force")),
    LIVARAG = factor(LIVARAG, labels = c("Homeless", "Dependent", "Independent")),
    ARRESTS_D = factor(ARRESTS_D, labels = c("0", "1", "2+")),
    SERVICES = factor(SERVICES, labels = c("Detox 24-hour outpatient", "Rehab/residential, <=30 days",
                                           "Rehab/residential, >30 days", "Ambulatory, intensive outpatient", "Ambulatory, non-intensive outpatient", "Ambulatory, detoxification")),
    DAYWAIT = factor(DAYWAIT, labels = c("0", "1-7", "8-14", "15-30", "31+")),
    REASON = factor(REASON, labels = c("Treatment completed", "Dropped out of treatment", "Terminated by facility", "Transferred", "Incarcerated", "Death", "Other")),
    PSOURCE = factor(PSOURCE, labels = c("Individual", "Alcohol/drug use care provider", "Other health care provider", "School", "Employer/EAP", "Other community", "Court")),
    NOPRIOR = factor(NOPRIOR, labels = c("0", "1+")),
    PSYPROB = factor(as.numeric(PSYPROB == 1)),
    FREQ_ATND_SELF_HELP = factor(FREQ_ATND_SELF_HELP, labels = c("0", "1-3", "4-7", "8-30")),
    LOS = factor(LOS, labels = c(seq(from = 1, to = 30), "31-45", "46-60", "61-90", "91-120", "121-180", "181-365", ">365"))
    )
tedsd_MI2017_los <- tedsd_MI2017_los %>% mutate_if(is.double, as.factor)
```

```{r}
percents_graphs_tables(~LOS|AGE, "Length of stay of opioid users by age at admission", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|GENDER, "Length of stay of opioid users by gender", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|RACE, "Length of stay of opioid users by race", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|MARSTAT, "Length of stay of opioid users by marital status", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|EDUC, "Length of stay of opioid users by education level", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|EMPLOY, "Length of stay of opioid users by employment", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|LIVARAG, "Length of stay of opioid users by living arrangement at admission", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|ARRESTS_D, "Length of stay of opioid users by arrests in the 30 days prior to discharge", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|SERVICES, "Length of stay of opioid users by services", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|DAYWAIT, "Length of stay of opioid users by days waiting before admission", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|REASON, "Length of stay of opioid users by reason for discharge", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|PSOURCE, "Length of stay of opioid users by referral source", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|NOPRIOR, "Length of stay of opioid users by number of prior admissions", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|PSYPROB, "Length of stay of opioid users by presence of co-occuring mental and substance use disorders", tedsd_MI2017_los, horiz=TRUE)
percents_graphs_tables(~LOS|FREQ_ATND_SELF_HELP, "Length of stay of opioid users by frequency of attending self-help programs in the past month", tedsd_MI2017_los, horiz=TRUE)
```
