---
title: "Models/Methods/Code Supplement"
author: "Catherine DeJager, Sarah Gilbert, and Kaitlyn Westra"
date: 'November 22, 2019'
output: 
  pdf_document:
    fig_height: 2.2
    fig_width: 4
  html_document:
    fig_height: 2.2
    fig_width: 4
  word_document:
    fig_height: 2.2
    fig_width: 4
---

```{r, setup, include = FALSE}
# load packages that are going to be used
require(tidyverse)   # this loads mosaic, ggformula, etc. too
require(ggformula)
require(mosaic)
require(plotly)
require(car)
require(glmmTMB)
require(MuMIn)

# Some customization.  You can alter or delete as desired (if you know what you are doing).

theme_set(theme_bw(base_size=12))     # change theme for ggplot2/ggformula

knitr::opts_chunk$set(
  echo = TRUE,      # for homework, always show R code (this is the default)
  tidy = FALSE,     # display code as typed (rather than reformatted)
  size = "small",   # slightly smaller font for code
  message = FALSE, warning = FALSE) # don't print warnings or messages in compiled document. So you MUST check them in RStudio!
```


### Read in Data

```{r}
load(file = "data/tedsd_2017.RData")
```

```{r}
tedsd_MI2017 <- filter(tedsd_puf_2017, STFIPS == 26)
tedsd_MI2017 <- tedsd_MI2017 %>%
  mutate(rowgroups = gl(n=(nrow(tedsd_MI2017)/250),
                        k = 250))
```

```{r}
tedsd_MI2017 <- tedsd_MI2017 %>% 
  mutate(OPIFLG = HERFLG==1 | METHFLG==1 | OPSYNFLG==1,
         HER1 = SUB1==5, HER2 = SUB2==5, HER3 = SUB3==5, METH1 = SUB1==6, METH2 = SUB2==6,
         METH3 = SUB3==6, OPSYN1 = SUB1==7, OPSYN2 = SUB1==7, OPSYN3 = SUB3==7)
tedsd_MI2017 <- tedsd_MI2017 %>% mutate(OPI1 = HER1|METH1|OPSYN1, OPI2 =
                                          HER2|METH2|OPSYN2, OPI3 = HER3|METH3|OPSYN3)
tedsd_MI2017 <- tedsd_MI2017 %>% mutate(MULTIOPI = (OPI1 & OPI2) | (OPI1 & OPI3) | 
                                          (OPI2 & OPI3))
```

```{r}
remove_nas <- function(df) {
  M <- as.matrix(df)
  M[M == -9] <- NA
  result <- as.data.frame(M)
  result <- result %>% na.omit()
}
```

```{r}
tedsd_MI2017_notna <- tedsd_MI2017 %>% select(HERFLG, METHFLG, OPSYNFLG, OPIFLG, HER1,
      HER2, HER3, METH1, METH2, METH3, OPSYN1, OPSYN2, OPSYN3, OPI1, OPI2, OPI3, MULTIOPI,
      AGE, GENDER, RACE, MARSTAT, EDUC, EMPLOY, LIVARAG, ARRESTS, NOPRIOR, PSYPROB,
      FREQ_ATND_SELF_HELP, LOS, PSOURCE, rowgroups)
tedsd_MI2017_notna <- tedsd_MI2017_notna %>% remove_nas()
```

```{r}
tedsd_MI2017_notna <- tedsd_MI2017_notna %>% mutate(
    AGE = factor(AGE, labels = c("12-14", "15-17", "18-20", "21-24", "25-29", "30-34",
                                 "35-39", "40-44", "45-49", "50-54", "55-64", "65+")),
    GENDER = factor(as.numeric(GENDER==2.0), labels = c("M", "F")),
    RACE = factor(RACE, labels = c("AK Native", "American Indian", "Black", "White",
                              "Asian", "Other", "Multiple", "HI/other pacific islander")),
    MARSTAT = factor(MARSTAT, labels = c("Never", "Married", "Separated",
                                         "Divorced/widowed")),
    EDUC = factor(EDUC, labels = c("<=8", "9-11", "12 or GED", "13-15", "16+")),
    EMPLOY = factor(EMPLOY, labels = c("Full-time", "Part-time", "Unemployed", 
                                       "Not in labor force")),
    LIVARAG = factor(LIVARAG, labels = c("Homeless", "Dependent", "Independent")) %>% relevel(ref = "Independent"),
    ARRESTS = factor(ARRESTS, labels = c("0", "1", "2+")),
    NOPRIOR = factor(NOPRIOR, labels = c("0", "1+")),
    PSYPROB = factor(as.numeric(PSYPROB == 1)),
    FREQ_ATND_SELF_HELP = factor(FREQ_ATND_SELF_HELP, labels = c("0", "1-3", "4-7",
                                                                 "8-30")),
    LOS = factor(LOS, labels = c(seq(from = 1, to = 30), "31-45", "46-60", "61-90",
                                 "91-120", "121-180", "181-365", ">365")),
    PSOURCE = factor(PSOURCE, labels = c("Individual", "Alcohol/drug use care provider",
      "Other health care provider", "School", "Employer/EAP", "Other community", "Court"))
    )
tedsd_MI2017_notna <- tedsd_MI2017_notna %>% mutate_if(is.double, as.factor)
```

```{r}
tedsd_MI2017_opi <- tedsd_MI2017_notna %>% filter(OPIFLG == TRUE)
head(tedsd_MI2017_opi)
```

```{r}
percents_graphs_tables <- function(form, ttl, my_df, horiz=FALSE) {
  if (horiz) {
    print(gf_percentsh(form, title=ttl, data=my_df))
  }
  else {
    print(gf_percents(form, title=ttl, data=my_df))
  }
  tally(form, format='percent', data=my_df) #%>% pander()
}
```



```{r}
predict_model <- function(model,predictor){
  orig_dat <- model@frame
  fixed_vals <- get_fixed(orig_dat[,(2:ncol(orig_dat))])
  new_dat <- get_new_data(orig_dat, predictor = predictor, fixed_vals)
  return(predict(model, newdata = new_dat,
                 type = 'response', allow.new.levels = TRUE))
}
```



### Models

#### Question 1: What variables predict whether a person entering treatment uses opioids or not?

We'll use a Logistic Regression here: 

(OPIFLG ~ GENDER + MARSTAT + GENDER\*MARSTAT + GENDER\*EMPLOY + LIVARAG + LIVARAG\*EDUC + LIVARAG*EMPLOY + NOPRIOR + FREQ_ATND_SELF_HELP)

EDUC = education, LIVARAG = living arrangement, NOPRIOR = whether the person has been admitted before, FREQ_ATND_SELF_HELP = frequency of attending self-help in the past 30 days

```{r, fig.height=4, fig.width=5}
opiModel <- glmmTMB(OPIFLG ~ GENDER + MARSTAT + GENDER*MARSTAT + GENDER*EMPLOY + LIVARAG + LIVARAG*EDUC + 
                  LIVARAG*EMPLOY + NOPRIOR + FREQ_ATND_SELF_HELP + (1|rowgroups), family=binomial(link="logit"), data=tedsd_MI2017_notna)
summary(opiModel)
acf(resid(opiModel),lag.max=1000)
```

This ACF was the best we could do. It's not *super* great, but it's a lot better than what we started with, indicating that our random effect variable (rowgroup) helped with the independence of residuals.

```{r}
Anova(opiModel)
```

##### The important predictors for predicting people admitted for opioid use are:
Risk factors: homeless, admitted previously, female and divorced/widowed, female, self help 4-7 times in the past month.
Mitigation factors: divorced/widowed, not in labor force, 9-11 years or 12/GED or 16+ years education, self help 8-30 times in the past month, unemployed and homeless, not in labor force and homeless.

Interactions occured between the variables:
Gender & Marital Status, Gender & Employement Status, Living Arrangement & Education, Living Arrangement & Employment Status, whether they had been admitted previously, and Frequency of Attending Self Help.


##### Prediction Plots:
Ideally, we would've liked to be able to show prediction plots for: gender, marstat, and noprior. However, because parametric bootstrapping was not working as expected, we were unable to do so.






#### Question 2: What variables predict whether a person entering treatment uses more than one kind of opioid?

We're using a Logistic Regression here:

(MULTIOPI ~ HERFLG + METHFLG + OPSYNFLG + AGE + GENDER + NOPRIOR)

NOPRIOR = whether the person has been admitted before

```{r}
multiOpiBest <- glmmTMB(MULTIOPI ~ HERFLG + METHFLG + OPSYNFLG + AGE + GENDER +
                      NOPRIOR + (1|rowgroups), data = tedsd_MI2017_opi,
                     family = binomial(link = "logit"))
```

```{r, fig.height=4, fig.width=5}
summary(multiOpiBest)
acf(resid(multiOpiBest))
```

Again, this ACF was the best we could do. It's a lot better than what we started with, indicating that our random effect variable (rowgroup) helped with the independence of residuals.


```{r}
Anova(multiOpiBest)
```


##### The important predictors for predicting people admitted for using multiple opioids are:
Risk factors: other opiates & synthetics, non-prescription methadone, 18 years or older, female.
Mitigation factors: heroin, admitted previously.

(No interactions occured between any predictors.)

##### Prediction Plots:

Ideally, we would've liked to be able to show prediction plots for: herflg, methflg, opsynflg. However, because parametric bootstrapping was not working as expected, we were unable to do so.







#### Question 3: What variables predict the length of stay for a person who enters treatment for opioid use? 

We're a bit confused because the response variable (Length of Stay) is similar to count data, except it has a some categories that have more than one day (including the category for more than 365 days). We don't know how to do multinomial regression. Is there a way we can coerce this into count data or should we pursue learning how to do multinomial regression? (See below:)

```{r}
tedsd_MI2017_los <- tedsd_MI2017 %>% select(HERFLG, METHFLG, OPSYNFLG, OPIFLG, HER1, HER2,
      HER3, METH1, METH2, METH3, OPSYN1, OPSYN2, OPSYN3, OPI1, OPI2, OPI3, MULTIOPI, AGE,
      GENDER, RACE, MARSTAT, EDUC, EMPLOY, LIVARAG, ARRESTS_D, SERVICES, METHUSE, DAYWAIT,
      REASON, PSOURCE, NOPRIOR, PSYPROB, FREQ_ATND_SELF_HELP, LOS, rowgroups) %>% filter(OPIFLG == 1)
tedsd_MI2017_los <- tedsd_MI2017_los %>% remove_nas()
```

```{r}
tedsd_MI2017_los <- tedsd_MI2017_los %>% mutate(LONGSTAY = as.numeric(LOS) > 13)
```

```{r}
tedsd_MI2017_los <- tedsd_MI2017_los %>% mutate(
    AGE = factor(AGE, labels = c("15-17", "18-20", "21-24", "25-29", "30-34", "35-39",
                                 "40-44", "45-49", "50-54", "55-64", "65+")),
    EDUC = factor(EDUC, labels = c("<=8", "9-11", "12 or GED", "13-15", "16+")),
    EMPLOY = factor(EMPLOY, labels = c("Full-time", "Part-time", "Unemployed", 
                                       "Not in labor force")),
    PSOURCE = factor(PSOURCE, labels = c("Individual", "Alcohol/drug use care provider",
        "Other health care provider", "School", "Employer/EAP", "Other community",
        "Court")),
    FREQ_ATND_SELF_HELP = factor(FREQ_ATND_SELF_HELP, labels = c("0", "1-3", "4-7",
                                                                 "8-30")),
    LOS = factor(LOS, labels = c(seq(from = 1, to = 30), "31-45", "46-60", "61-90",
                                 "91-120", "121-180", "181-365", ">365"))
    )
```

```{r, fig.height=5, fig.width=7}
percents_graphs_tables(~LOS, "Length of stay (in days) of opioid users", tedsd_MI2017_los,
                       horiz=TRUE)
```

```{r}
tally(~PSOURCE, data=tedsd_MI2017_los)
```

```{r}
los_lrm <- glmmTMB(LONGSTAY ~ AGE + EDUC + EMPLOY + FREQ_ATND_SELF_HELP + PSOURCE + OPSYNFLG + (1|rowgroups), data=tedsd_MI2017_los, family=binomial(link = "logit"))
```

```{r}
summary(los_lrm)
```

```{r, fig.height=4, fig.width=5}
acf(resid(los_lrm))
```

Again, this ACF was the best we could do. It's a lot better than what we started with, indicating that our random effect variable (rowgroup) helped with the independence of residuals.


```{r}
Anova(los_lrm)
```


##### The important predictors for predicting stays of over 2 weeks are:

Longer stay: referred by employer/EAP, referred by court, referred by alcohol/drug use care provider, attended self-help in the past month, referred by other community, 16+ years of education, other opioids & synthetics
Shorter stay: 18 years or older, not in labor force


##### Prediction Plots:

Ideally, we would've liked to be able to show prediction plots for: freq_atnd_self_help, psource, opsynflg. However, because parametric bootstrapping was not working as expected, we were unable to do so.

